---
title: "Customer Segmentation"
author: "Akanbi Abdullahi"
date: "2024-08-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```


```{r}
library(dplyr)
library(fastDummies)
library(factoextra)
library(ggpubr)
library(tidyr)
library(ggplot2)
library(plotrix)
library(cluster) 
```



#getting the data ready
```{r}
data <- read.csv("customer_data.csv", header=TRUE)
data
summary(data)
```

#removing the first column; ID
```{r}
data <- data[,2:9]
data
```

#checking for the number of profession in the dataset by checking unique values.
```{r}
unique(data$Profession)
```

#transforming empty values to "NA"
```{r}
data[data == ""] <- NA
data <- data[complete.cases(data),]
```

#building the dataset
```{r}
dataset <- data %>% select_if(is.numeric)
character <- data %>% select_if(is.character)
```

#transform character into numeric
```{r}
character <- dummy_cols(character, remove_most_frequent_dummy = TRUE)

```

#remove the characters in the data
```{r}
dataset <- cbind(dataset, character[,6:18]) 
```
#scaling the dataset
```{r}
dataset[, 1: 16] <- scale(dataset[, 1: 16])
```

#Determining and vizualizing the optimal number of clusters
```{r}
fviz_nbclust(dataset, kmeans, method = "wss") + labs(subtitle = "Elbow Method")
```

#Clusters
```{r}
#clusters <-kmeans(dataset,centers = 6, iter.max = 10)
#clusters$centers
set.seed(123)
k <- 6  # Based on elbow method
kmeans_model <- kmeans(dataset, centers = k, nstart = 25)
kmeans_model$centers
```

```{r}
#Calculating the silhouette score for K-Means clustering
silhouette_kmeans <- silhouette(kmeans_model$cluster, dist(dataset))

#Getting the average silhouette score
avg_silhouette_kmeans <- mean(silhouette_kmeans[, 3])
```


```{r}
#Visualizing the silhouette plot
fviz_silhouette(silhouette_kmeans) +
  ggtitle(paste("Silhouette Plot for K-Means (K =", k, ") - Average Score: ", round(avg_silhouette_kmeans, 3)))
```

```{r}
# Output the average silhouette score
print(paste("Average Silhouette Score for K =", k, ":", avg_silhouette_kmeans))
```


#writing these clusters into a csv file for analysis
```{r}
write.csv(clusters$centers, file = "clusters.csv")
```

#adding the clusters to the customer list
```{r}
dataset <- cbind(dataset , clusters$cluster)
```


#visualizing the clusters
```{r}
 set.seed(1)
```


```{r}
pca <- prcomp(dataset, center = TRUE, scale. = TRUE)
# Ensure clusters$cluster contains the cluster assignments
pca_data <- data.frame(pca$x, cluster = as.factor(clusters$cluster))
```

```{r}
# Visualizing the clusters
# Plot PCA with cluster assignments
ggplot(pca_data, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point() +
  labs(x = "Principal Component 1", y = "Principal Component 2", color = "Cluster") +
  ggtitle("PCA Plot with K-Means Clusters")
```


#Gender visualization of the dataset
```{r}
a <- table(data$Gender)
barplot(a, main = "Using BarPlot to display Gender Comparison",
        ylab = "Count", xlab = "Gender",
        col = rainbow(length(a)), # Use rainbow() to get a vector of colors
        legend = rownames(a))
```

```{r}
pct=round(a/sum(a)*100) 
lbs=paste(c("Female","Male"),"",pct,"%",sep=" ")
pie3D(a,labels=lbs, main="Pie Chart Depicting Ratio of Female and Male")
```
#Age distribution of the dataset
```{r}
# Defining the age classes and their corresponding colors
age_classes <- c("0-20", "21-40", "41-60", "61-80", "81-100")
age_colors <- c("blue", "green", "orange", "red", "purple")

# Creating the histogram with custom age classes and colors
hist(data$Age,
     breaks = c(0, 20, 40, 60, 80, 100),
     col = age_colors,
     main = "Histogram to Show Count of Age Class",
     xlab = "Age Class",
     ylab = "Frequency",
     labels = TRUE)

# Adding a legend
legend("topright",
       legend = age_classes,
       fill = age_colors,
       title = "Age Classes")
```





